Jane, John … Leslie? A Historical Method for Algorithmic Gender Prediction

### 2015-08-15 article.Rmd
Revise description of datasets

The possibilities for inferring gender from names depends on two things. First, it needs a suitable (and suitably large) dataset for the time period and region under study. Unsurprisingly such datasets are almost always gathered in the first instance by governments, though their compilation and digitization may be undertaken by scholarly researchers. Second, it depends on a suitable algorithm for estimating the proportion of male and female names for a given year or range of years, since often a person cannot be associated with an exact date. It is especially important that the algorithm take into account any biases in the data to formulate more accurate predictions. Development of the R package has had two primary aims. The first is to abstract the predictive algorithm to the simplest possible form so that it is usable for a wide range of historical problems rather than depending on the format of any particular data set. The second has been to provide as many datasets as possible in order to localize the predictions to particular times and places. To that end, [Benjamin Schmidt](http://benschmidt.org/) contributed a dataset of US Census Data and accompanying code, as well as provided a significant correction to predictions based on Social Security Administration data.

The Gender package currently uses two datasets which make it suitable for studying the United States from the first federal census in 1790 onwards.^[The authors of the package are both American historians, which explains why the initial release of the package focuses on that country. We have plans, however, to extend the underlying data to six other European and North American countries included in the [North Atlantic Population Project](https://www.nappdata.org/napp/) for the nineteenth century. We also welcome contributions of datasets for other regions and times from other scholars.] The first dataset contains names of applicants for Social Security and is available from [Data.gov](https://www.data.gov/). The second dataset contains names gathered in the decennial censuses and is available from the IPUMS-USA (Integrated Public Use Microdata Series) service from the Minnesota Population Center at the University of Minnesota.

The Social Security Administration (SSA) [Baby Names](http://catalog.data.gov/dataset/baby-names-from-social-security-card-applications-national-level-data) dataset was created as a result of the Social Security Act of 1935 during the New Deal.^[The Social Security Administration also provides a related dataset which contains the names of applicants broken down by state and year since 1910 [Social Security Administration 2014-b].] This dataset contains a list of first names along with how many times each name was assigned to each gender in every year beginning with 1880 and ending with 2012. The SSA list includes any name which was used more than five times in a given year, thereby capturing all but the most infrequently used names from each year. The description "baby names" provided by the SSA is a serious misnomer. When Social Security became available during the New Deal, its first beneficiaries were adults past or near retirement age. The dataset goes back to 1880, the birth year for a 55 year-old adult when Social Security was enacted. Even after 1935, registration at birth for Social Security was not mandatory until 1986. As we will demonstrate below, the way in which the data was gathered requires an adjustment to our predictions of gender.^[Shane Landrum has studied the history of birth registration in the United States [Landrum 2014]. It is well-known among American historians that the Social Security Act was written to exclude farmworkers as a means of also excluding African Americans. In examining the dataset we have not found any reason to believe that this policy noticably skewed the prediction of gender from names.]

The IPUMS-USA dataset, contributed by Benjamin Schmidt, contains records from the United States decennial census from 1790 to 1930. This dataset includes the birth year and numbers of males and females under the age of 62 for all the years in that range. This data has been aggregated by IPUMS at the University of Minnesota and is released as a sample of the total census data. Unlike the SSA dataset, which includes a 100% sample for every name reported to the Social Security Administration and used more than five times, the IPUMS data contains 5% or 10% samples of names from the total census data. Because the `gender()` function relies on proportions of uses of names, rather than raw counts of people with the names, the sampling does not diminish the reliability of the function's predictions [Ruggles et al. 2010].

These two SSA and IPUMS datasets (and any future datasets to be added to the package) thus contain a simple table of data.^[The format of this data departs from the more generally desirable "tidy data" framework proposed by Hadley Wickham for performance reasons. By keeping the data in a "wider" format instead of a tidy format as defined by Wickham, the function is able to make its predictions faster [Wickham 2014b].] The columns contain the `name`, `year`, and number of `female` and `male` uses of that name in a particular year.

>>>>

Inferring gender from names depends on two things. First, it requires a suitable (and suitably large) dataset for the time period and region under study. Unsurprisingly such datasets are almost always gathered in the first instance by governments, though their compilation and digitization may be undertaken by scholarly researchers. Second, it depends on a suitable algorithm for estimating the proportion of male and female names for a given year or range of years, since often a person cannot be associated with an exact date. It is especially important that the algorithm take into account any biases in the data to formulate more accurate predictions. Development of the R package has had two primary aims. The first is to abstract the predictive algorithm to the simplest possible form so that it is usable for a wide range of historical problems rather than depending on the format of any particular data set. The second has been to provide as many datasets as possible in order to localize the predictions to particular times and places.

The Gender package currently uses several datasets which make it suitable for studying the United States from the first federal census in 1790 onwards.] The first dataset contains names of applicants for Social Security and is available from [Data.gov](https://www.data.gov/). The second dataset contains names gathered in the decennial censuses and is available from the IPUMS-USA (Integrated Public Use Microdata Series) service from the Minnesota Population Center at the University of Minnesota.

The Social Security Administration (SSA) [Baby Names](http://catalog.data.gov/dataset/baby-names-from-social-security-card-applications-national-level-data) dataset was created as a result of the Social Security Act of 1935 during the New Deal.^[The Social Security Administration also provides a related dataset which contains the names of applicants broken down by state and year since 1910 [Social Security Administration 2014-b].] This dataset contains a list of first names along with how many times each name was assigned to each gender in every year beginning with 1880 and ending with 2012. The SSA list includes any name which was used more than five times in a given year, thereby capturing all but the most infrequently used names from each year. The description "baby names" provided by the SSA is a serious misnomer. When Social Security became available during the New Deal, its first beneficiaries were adults past or near retirement age. The dataset goes back to 1880, the birth year for a 55 year-old adult when Social Security was enacted. Even after 1935, registration at birth for Social Security was not mandatory until 1986. As we will demonstrate below, the way in which the data was gathered requires an adjustment to our predictions of gender.^[Shane Landrum has studied the history of birth registration in the United States [Landrum 2014]. It is well-known among American historians that the Social Security Act was written to exclude farmworkers as a means of also excluding African Americans. In examining the dataset we have not found any reason to believe that this policy noticeably skewed the prediction of gender from names.]

The IPUMS-USA dataset, contributed by [Benjamin Schmidt](http://benschmidt.org/), contains records from the United States decennial census from 1790 to 1930. This dataset includes the birth year and numbers of males and females under the age of 62 for all the years in that range. This data has been aggregated by [IPUMS](https://usa.ipums.org/usa/) at the University of Minnesota and is released as a sample of the total census data. Unlike the SSA dataset, which includes a 100% sample for every name reported to the Social Security Administration and used more than five times, the IPUMS data contains 5% or 10% samples of names from the total census data. Because the `gender()` function relies on proportions of uses of names, rather than raw counts of people with the names, the sampling does not diminish the reliability of the function's predictions [Ruggles et al. 2010].

The next release of the Gender package will include a dataset from the [North Atlantic Population Project](https://www.nappdata.org/napp/). This dataset will include names from Canada, Great Britain, Germany, Iceland, Norway, and Sweden from 1801 to 1910.^[The North Atlantic Population Project, which is also project of the Minnesota Population Center, also includes data for the United States of America, but we have excluded it since IPUMS-USA and the SSA dataset is more suitable when focusing on that country alone.] The SSA, IPUMS, and NAPP datasets (and any future datasets to be added to the package) all have the same simple tabular format.^[The format of this data departs from the more generally desirable "tidy data" framework proposed by Hadley Wickham for performance reasons. By keeping the data in a "wider" format instead of a tidy format as defined by Wickham, the function is able to make its predictions faster [Wickham 2014b].] The columns contain the `name`, `year`, and number of `female` and `male` uses of that name in a particular year. This simple format makes it possible to extend the package to include any place and time period for which there is suitable data.^[We welcome such contributions at the [project's GitHub repository](https://github.com/ropensci/gender).]
